How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-04-29 doc/paper.tex
paper ready for submission process

\begin{abstract}
 A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) For each group we examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate state) the improvement in predictive power over models with constant effects and no historical covariates is very minor. This model has an approximately 78\% median probability of correctly ranking the relative extinction risk any two randomly selected species. It performs nearly as well at out-of-sample prediction (70-80\%), implying that determinants of extinction risk have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

We used a discrete-time survival modelling framework to estimate how well we can predict extinction risk at one million year time scales. At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}. We considered four different models involving different permutations of covariate effects (fixed or time-varying) and historical covariates: covariate effects constant over time and no historical covariates included (Model C), covariate effects allowed to vary over time but no historical covariates included (Model V), covariate effects constant over time and historical covariates included (Model CP), and covariate effects allowed to vary over time and historical covariates are included (Model VP). The C and P models atempt to predict based only on present state, wheras the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.

We always included species age at time of observation (.e. observed prior duration) as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.

@@

k-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold. Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold. The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

@@

We find that all of our models are expected to correctly forecasting the rank order in extinction probabilities for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic. A extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

@@

The relative quality and consistency of our models out-of-sample forecasting performance is encouraging given that its estimates are based on very limited biological and environmental information about the taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and taxonomy. The principal reason we were not able to include more biological information in our models is that we lack suitably detailed life history or ecological information for many marine micro- and nannoplankton. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Further analyses including this type of information and focused more narrowly on the foraminifera may be informative.

@@

In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting the relative extinction risk of two randomly selected species based on a small number of taxonomic, geographic, and historical predictors. Considering that conservation decisions are made based on a continuum of risk, from most to least, this means that our results are in the same language as how conservation resources are allocated. These results present an encouraging baseline for using the fossil record to predict future extinction. This suggests, that all else being equal, extending this model to include the information in modern risk assessments might actually do decently well at forecasting future extinctions.

@@

\section{Supplement to Materials and Methods}

>>>>

\begin{abstract}
 A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate) the improvement in predictive power over models with constant effects and no historical covariates is very minor. Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample pair, implying that determinants of extinction have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

We used a discrete-time survival modelling framework to estimate how well we can predict extinction risk at one million year time scales. At its core, our model is a multilevel logistic regression with taxon age in millions of years as a varying intercept \citep{Tutz2016}. We considered four different models involving different permutations of covariate effects (fixed or time-varying) and historical covariates: covariate effects constant over time and no historical covariates included (Model C), covariate effects allowed to vary over time but no historical covariates included (Model V), covariate effects constant over time and historical covariates included (Model CP), and covariate effects allowed to vary over time and historical covariates are included (Model VP). The C and P models attempt to predict based only on present state, whereas the CP and VP models allow for the possibility of non-Markovian behaviour by including change in state from the previous time increment.

We always included species age at time of observation (i.e. observed prior duration) as a non-nested varying intercept term. This factor may or may not contribute to differences in species extinction risk over time \citep{Smits2015,Finnegan2008,Ezard2012,VanValen1973,Liow2011,Crampton2016}, but its inclusion in our model is critical to its nature as a survival model \cite{Tutz2016}.

@@

\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold. Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold. The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

@@

We find that all of our models are expected to correctly forecasting the rank order of extinction probability for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic. A extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

@@

The relative quality and consistency of our models out-of-sample forecasting performance is encouraging given that its estimates are based on very limited biological and environmental information about the taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and taxonomy. The principal reason we were not able to include more biological information in our models is that we lack suitably detailed life history or ecological information for many marine micro- and nannoplankton. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.

@@

@@

\setcounter{page}{1}
\section{Supplement to Materials and Methods}
